$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-12-14 paper/paper.tex
Fix cursive in definitions, lemmas and theorems

% Continue examples later on
\usepackage{thmtools}
\renewcommand\thmcontinues[1]{Continued}

% Do not use italics in definitions and theorems
% \theoremstyle{definition}
% \newtheorem{definition}{Definition}
% \newtheorem{theorem}{Theorem}
% \newtheorem{lemma}{Lemma}
% % \newtheorem{example}{Example}
% \renewcommand{\definitionautorefname}{Definition}%
% \renewcommand{\theoremautorefname}{Theorem}%
% \renewcommand{\lemmaautorefname}{Lemma}%
% \renewcommand{\sectionautorefname}{\S}%
% \renewcommand{\subsectionautorefname}{\S}%
% \renewcommand{\exampleautorefname}{Example}

@@

\begin{abstract}
  \todo{the only references here are to session types}
  \todo{fix acknowledgments}
  \todo{fix lemma, definition, theorem, example envs}
  The \picalc{} is a computational model for communication and concurrency.
  The \emph{linear} \picalc{} is a typed version of the \picalc{} where channels must be used exactly once.
  It is an underlying theoretical and practical framework on top of which more advanced types and theories can be built, such as \emph{session types} \cite{H93,THK94,HVK98}.

@@

\begin{definition}
  \label{def:unused}
  A variable $i$ is unused in $P$ ($\Unused_i \; P$) if it appears in none of its inputs and none of its outputs.
  \todo{should we have this as a figure instead?}

@@

  The type $i \type{\not\equiv} x$ unfolds to the negation of propositional equality on \textsc{Var}, i.e. $i \type{\equiv} x \to \type{\bot}$.
\end{definition}

@@

\begin{lemma}
  \label{lm:substitution-unused}
  For every variables $i$ and $j$, if $i \type{\not\equiv} j$ then $\Unused_i \allowbreak (\subst{P}{j}{i})$.
\end{lemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{definition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$ that is \emph{partial} (as not any two multiplicities can be combined), \emph{deterministic} and \emph{cancellative} (to aid equational reasoning), \emph{associative} and \emph{commutative} (following directly from subject congruence for parallel composition), and in which the leftovers can be \emph{computed} (as to automatically update the usage context every time input and output occurs).
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal} (so that new resources cannot arbitrarily spring into life).
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} defines such an algebra as a record $\Algebra_C$ on a carrier $C$.

  \begin{figure}[h]
  \begin{equation*}
  \begin{aligned}
    &\zero                   &:{} &                 &        & C \\
    &\one                    &:{} &                 &        & C \\
    &\op{\_}{\_}{\_}         &:{} &                 &        & C \to C \to C \to \Set \\
    &\func{\cdot-compute^r}  &:{} &\forall x y       & \to \; & \type{DEC} \; (\type{\exists} z  \; (\op{x}{y}{z})) \\
    &\func{\cdot-unique}     &:{} &\forall x x' y z  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\func{\cdot-unique^l}   &:{} &\forall x y y' z  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\func{\cdot-min^l}      &:{} &\forall y z        & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\func{\cdot-id^l}       &:{} &\forall x         & \to \; & \op{x}{\zero}{x} \\
    &\func{\cdot-comm}       &:{} &\forall x y z     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\func{\cdot-assoc}      &:{} &\forall x y z u v & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \\
    &                        &    &                  &        & \type{\exists} w  \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation*}
  \caption{
    Usage algebra $\Algebra_C$ on a carrier $C$.
    We use $\forall$ for universal quantification.
    The dependent product $\type{\exists}$ uses the value of its first argument in the type of its second.
    The type $\type{DEC} \; P$ is a witness of either $P$ or $P \to \type{\bot}$, where $\type{\bot}$ is the empty type with no constructors.
  }
  \label{fig:multiplicities}
  \end{figure}
\end{definition}

@@

We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system with leftovers (\autoref{leftover-typing}).
\begin{definition}[Indexed set of usage algebras]
  An \emph{indexed set of usage algebras} is a type $\Idx$ of indices that is nonempty ($\type{\exists IDX}$) together with an interpretation $\Usage$ of indices into types, and an interpretation $\type{ALGEBRAS}$ of indices into usage algebras of the corresponding type (\autoref{fig:indexed-multiplicities}).
  \begin{figure}[h]
    \begin{equation*}
      \begin{aligned}
        &\Idx               &:{} &\Set \\
        &\type{\exists IDX} &:{} &\Idx \\
        &\Usage             &:{} &\Idx \to \Set \\
        &\type{ALGEBRAS}    &:{} &(idx : \Idx) \to \Algebra_{\Usage_{idx}}
      \end{aligned}
    \end{equation*}
    \caption{Indexed set of usage algebras.}
    \label{fig:indexed-multiplicities}
  \end{figure}
\end{definition}

@@

\begin{definition}[\textsc{Type} and \textsc{PreCtx}: types and typing contexts]
  A \emph{type} is either a unit type ($\unit$), a base type ($\base{n}$) or a channel type ($\channel{t}{x}$) (first two rows of \autoref{fig:prectx}).

@@

A \emph{typing context} (last two rows of \autoref{fig:prectx}) is a length-indexed list of types, and is either empty $[]$ or the result of appending a type $t$ to an existing context $\gamma$.
\end{definition}

\begin{definition}[\textsc{Ctx}: usage contexts]
  A \emph{usage context} is a context $\Ctx_{idxs}$ of pairs of usage annotations that is indexed by a length-indexed context of indices $\Idxs_n$ (\autoref{fig:ctx}).

@@

    \inferrule
        {\Gamma : \Ctx_{idxs} \\ x : \Usage_{idx} ^{\func{2}}}
        {\Gamma \comma x : \Ctx_{idxs \comma idx}}
  \end{mathpar}
  \caption{Length-indexed context of carrier indices with a context of usage annotations on top.}
  \label{fig:ctx}
  \end{figure}
\end{definition}

@@

\begin{definition}[\textsc{VarRef}: typing variable references]
  \label{def:varref}

@@

  The base case $\constr{0}$ splits the usage annotation $x$ of type $\Usage_{idx}$ into $y$ and $z$ (the leftovers).
  Note that the remaining context $\Gamma$ is preserved unused as a leftover.
  This splitting $\opsquared{x}{y}{z}$ is as per the usage algebra provided by the developer for the index $idx$.
  We use the \emph{computability} of the monoidal relation to alleviate the user from the proof burden $\opsquared{x}{y}{z}$.
  The inductive case $\suc$ appends the type $t'$ to the typing context, and the usage annotation $x'$ to both the input and output usage contexts. 
\end{definition}

@@

\begin{definition}[\textsc{Types}: typing processes]
  \label{def:types}

@@

  The inaction process in rule $\PO$ does not change usage annotations.
  The scope restriction in rule $\new$ expects three arguments: the type $t$ of data being transmitted; the usage annotation $x$ of what is being transmitted; and the multiplicity $y$ given to the channel itself.
  This multiplicity $y$ is used for both input and output, so that they are balanced.
  The continuation process $P$ is provided with the new channel with usage annotation $y \comma y$, which it must completely exhaust.
%
  The input process in rule $\recv{}{}$ requires a channel $chan_i$ at index $i$ with usage $\li$ available, such that data with type $t$ and usage $x$ can be sent over it.
  Note that the index $i$ is used in the syntax of the typed process.
  We use the leftovers $\Xi$ to type the continuation process, which is also provided with the received element --- of type $t$ and multiplicity $x$ --- at index $\constr{0}$.
  The received element $x$ must be completely exhausted by the continuation process.
%
  Similarly to input, the output process in rule $\send{}{}$ requires a channel $chan_i$ at index $i$ with usage $\lo$ available, such that data with type $t$ and usage $x$ can be sent over it.
  We use the leftover context $\Delta$ to type the transmitted data, which needs an element $loc_j$ at index $j$ with type $t$ and usage $x$, as per the type of the channel $chan_i$.
  The leftovers $\Xi$ are used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.
%
  Parallel composition in rule $\comp{}{}$ uses the leftovers of the left-hand process to type the right-hand process.
  By keeping track in the typing derivation of $P$ of the resources $P$ uses, we can use them to type $Q$ and save the user from having to provide a top-down proof of context split.
\end{definition}

@@

\begin{theorem}[Framing]
  \label{thm:framing}
  Let $P$ be well typed in $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$.
  Let $\Delta$ be such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Let $\Gamma_r$ and $\Xi_r$ be arbitrary contexts such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{theorem}

\paragraph*{Weakening}
This property states that inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.
Let $\func{ins}_i$ insert an element into a context at position $i$ --- for simplicity, we use it both to insert types into typing contexts and usage annotations into usage contexts.
\begin{theorem}[Weakening]
  \label{thm:weakening}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, lifting every variable greater than or equal to $i$ in $P$ is well typed in
  $\types{\func{ins}_i \; t \; \gamma}{\func{ins}_i \; x \; \Gamma}{\func{lift}_i \; P}{\func{ins}_i \; x \; \Xi}$.
\end{theorem}

\paragraph*{Strengthening}
This property states that removing an unused variable preserves the well-typedness of a process.
Let $\func{del}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
\begin{theorem}[Strengthening]
  \label{thm:strengthening} 
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable not in $P$, such that $uP \; : \; \Unused_i \; P$.
  Then lowering every variable greater than $i$ in $P$ is well typed in $\types{\func{del}_i \; \gamma}{\func{del}_i \; \Gamma}{\func{lower}_i \; P \; uP}{\func{del}_i \; \Xi}$.
\end{theorem}

\paragraph*{Exchange}
This property states that the exchange of two variables preserves the well-typedness of a process.
We extend $\func{exchange}_i$  introduced in \autoref{structural-congruence} to exchange types in typing contexts and usage annotations in usage contexts.
\begin{theorem}[Exchange]
  \label{thm:exchange}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, $\types{\func{exchange}_i \; \gamma}{\func{exchange}_i \; \Gamma}{\allowbreak \func{exchange}_i \; P}{\func{exchange}_i \; \Xi}$.
\end{theorem}

@@

\begin{lemma}
  \label{lm:lower-lift}
  The function $\func{lower}_i \; P \; uP$ has an inverse $\func{lift}_i \; P$ that increments every $\textsc{Var}$ greater than or equal to $i$, such that $\func{lift}_i \; (\func{lower}_i \; P \; uP) \equiv P$.
\end{lemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{lemma}
  \label{lm:exchange-exchange}
  The function $\func{exchange}_i \; P$ is its own inverse: $\func{exchange}_i \; (\func{exchange}_i \; P) \equiv P$.
\end{lemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{lemma}
  \label{lm:types-unused}
  For all well-typed processes $\types{\gamma}{\Gamma}{P}{\Xi}$, if the variable $i$ is unused within $P$, then $\Gamma$ at $i$ is equivalent to $\Xi$ at $i$.
\end{lemma}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{theorem}[Subject congruence]
  \label{thm:subject-congruence}
  If $P \eq{} Q$ and $\types{\gamma}{\Gamma}{P}{\Xi}$, then $\types{\gamma}{\Gamma}{Q}{\Xi}$.
\end{theorem}

@@

\begin{theorem}[Generalised substitution]
  \label{thm:substitution-generalization}
  Let process $P$ be well typed in $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  Then, there must exist some $\Gamma$, $\Psi$, $\Gamma_j$ and $\Psi_j$ such that:
  \begin{multicols}{2}
  \begin{itemize}
    \item $\contains{\gamma}{\Gamma_i}{i}{t}{m}{\Gamma}$
    \item $\contains{\gamma}{\Gamma_j}{j}{t}{m}{\Gamma}$
    \item $\contains{\gamma}{\Psi_i}{i}{t}{n}{\Psi}$
    \item $\contains{\gamma}{\Psi_j}{j}{t}{n}{\Psi}$
  \end{itemize}
  \end{multicols}
  Let $\Gamma$ and $\Psi$ be such that $\opctx{\Gamma}{\Delta}{\Psi}$ for some $\Delta$.
  Let $\Delta$ at position $i$ have usage $\lz$, meaning all consumption from $m$ to $n$ must happen in $P$.
  Then substituting $i$ to $j$ in $P$ will be well typed in $\types{\gamma}{\Gamma_j}{\subst{P}{j}{i}}{\Psi_j}$.
\end{theorem}

@@

\begin{theorem}[Substitution]
  \label{thm:substitution}
  Let process $P$ be well typed in $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \lz}$.
  Let $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute the variable references to $\constr{0}$ in $P$ with $\suc j$ so that the result is well typed in $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{0}}}{\Xi \comma m}$.
\end{theorem}
\begin{proof}[Proof]
  For $\contains{\gamma}{\Gamma}{j}{t}{m}{\Theta}$ and $\types{\gamma \comma t}{\Theta \comma m}{P}{\Xi \comma \lz}$ for some $\Theta$, we use framing to derive them.
  Then, we use these to apply \autoref{thm:substitution-generalization}.
\end{proof}

@@

\begin{lemma}
  \label{lm:comm-capable}
  Every input usage context $\Gamma$ of a well-typed process $\types{\gamma}{\Gamma}{P}{\Delta}$ that reduces by communicating on a channel external (that is, $P \reduce{\constr{external} \; i} Q$ for some $Q$) has a multiplicity of at least $\lio$ at index $i$.
\end{lemma}

@@

\begin{theorem}[Subject reduction]
  \label{thm:subject-reduction}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$ and reduce such that $P \reduce{c} Q$.
  \begin{itemize}
    \item If $c$ is $\constr{internal}$, then $\types{\gamma}{\Gamma}{Q}{\Xi}$.
    \item If $c$ is $\constr{external} \; i$ and $\containsusage{\Gamma}{i}{\lio}{\Delta}$, then $\types{\gamma}{\Delta}{Q}{\Xi}$.
  \end{itemize}
\end{theorem}

>>>>

% Continue examples later on
% \usepackage{thmtools}
% \renewcommand\thmcontinues[1]{Continued}

% Do not use italics in definitions and theorems
\spnewtheorem{nidefinition}[definition]{Definition}{\bfseries}{}
\spnewtheorem{nilemma}[lemma]{Lemma}{\bfseries}{}
\spnewtheorem{nitheorem}[theorem]{Theorem}{\bfseries}{}

\renewcommand{\sectionautorefname}{\S}%
\renewcommand{\subsectionautorefname}{\S}%

@@

\begin{abstract}
  \todo{the only references here are to session types}
  \todo{fix acknowledgments}
  The \picalc{} is a computational model for communication and concurrency.
  The \emph{linear} \picalc{} is a typed version of the \picalc{} where channels must be used exactly once.
  It is an underlying theoretical and practical framework on top of which more advanced types and theories can be built, such as \emph{session types} \cite{H93,THK94,HVK98}.

@@

\begin{nidefinition}
  \label{def:unused}
  A variable $i$ is unused in $P$ ($\Unused_i \; P$) if it appears in none of its inputs and none of its outputs.
  \todo{should we have this as a figure instead?}

@@

  The type $i \type{\not\equiv} x$ unfolds to the negation of propositional equality on \textsc{Var}, i.e. $i \type{\equiv} x \to \type{\bot}$.
\end{nidefinition}

@@

\begin{nilemma}
  \label{lm:substitution-unused}
  For every variables $i$ and $j$, if $i \type{\not\equiv} j$ then $\Unused_i \allowbreak (\subst{P}{j}{i})$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{nidefinition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$ that is \emph{partial} (as not any two multiplicities can be combined), \emph{deterministic} and \emph{cancellative} (to aid equational reasoning), \emph{associative} and \emph{commutative} (following directly from subject congruence for parallel composition), and in which the leftovers can be \emph{computed} (as to automatically update the usage context every time input and output occurs).
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal} (so that new resources cannot arbitrarily spring into life).
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} defines such an algebra as a record $\Algebra_C$ on a carrier $C$.

  \begin{figure}[h]
  \begin{equation*}
  \begin{aligned}
    &\zero                   &:{} &                 &        & C \\
    &\one                    &:{} &                 &        & C \\
    &\op{\_}{\_}{\_}         &:{} &                 &        & C \to C \to C \to \Set \\
    &\func{\cdot-compute^r}  &:{} &\forall x y       & \to \; & \type{DEC} \; (\type{\exists} z  \; (\op{x}{y}{z})) \\
    &\func{\cdot-unique}     &:{} &\forall x x' y z  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\func{\cdot-unique^l}   &:{} &\forall x y y' z  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\func{\cdot-min^l}      &:{} &\forall y z        & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\func{\cdot-id^l}       &:{} &\forall x         & \to \; & \op{x}{\zero}{x} \\
    &\func{\cdot-comm}       &:{} &\forall x y z     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\func{\cdot-assoc}      &:{} &\forall x y z u v & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \\
    &                        &    &                  &        & \type{\exists} w  \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation*}
  \caption{
    Usage algebra $\Algebra_C$ on a carrier $C$.
    We use $\forall$ for universal quantification.
    The dependent product $\type{\exists}$ uses the value of its first argument in the type of its second.
    The type $\type{DEC} \; P$ is a witness of either $P$ or $P \to \type{\bot}$, where $\type{\bot}$ is the empty type with no constructors.
  }
  \label{fig:multiplicities}
  \end{figure}
\end{nidefinition}

@@

We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system with leftovers (\autoref{leftover-typing}).
\begin{nidefinition}[Indexed set of usage algebras]
  An \emph{indexed set of usage algebras} is a type $\Idx$ of indices that is nonempty ($\type{\exists IDX}$) together with an interpretation $\Usage$ of indices into types, and an interpretation $\type{ALGEBRAS}$ of indices into usage algebras of the corresponding type (\autoref{fig:indexed-multiplicities}).
  \begin{figure}[h]
    \begin{equation*}
      \begin{aligned}
        &\Idx               &:{} &\Set \\
        &\type{\exists IDX} &:{} &\Idx \\
        &\Usage             &:{} &\Idx \to \Set \\
        &\type{ALGEBRAS}    &:{} &(idx : \Idx) \to \Algebra_{\Usage_{idx}}
      \end{aligned}
    \end{equation*}
    \caption{Indexed set of usage algebras.}
    \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}

@@

\todo{definition captions reset textsc}
\begin{nidefinition}[\textsc{Type} and \textsc{PreCtx}: types and typing contexts]
  A \emph{type} is either a unit type ($\unit$), a base type ($\base{n}$) or a channel type ($\channel{t}{x}$) (first two rows of \autoref{fig:prectx}).

@@

A \emph{typing context} (last two rows of \autoref{fig:prectx}) is a length-indexed list of types, and is either empty $[]$ or the result of appending a type $t$ to an existing context $\gamma$.
\end{nidefinition}

\begin{nidefinition}[\textsc{Ctx}: usage contexts]
  A \emph{usage context} is a context $\Ctx_{idxs}$ of pairs of usage annotations that is indexed by a length-indexed context of indices $\Idxs_n$ (\autoref{fig:ctx}).

@@

    \inferrule
        {\Gamma : \Ctx_{idxs} \\ x : \Usage_{idx} ^{\func{2}}}
        {\Gamma \comma x : \Ctx_{idxs \comma idx}}
  \end{mathpar}
  \caption{Length-indexed context of carrier indices with a context of usage annotations on top.}
  \label{fig:ctx}
  \end{figure}
\end{nidefinition}

@@

\begin{nidefinition}[\textsc{VarRef}: typing variable references]
  \label{def:varref}

@@

  The base case $\constr{0}$ splits the usage annotation $x$ of type $\Usage_{idx}$ into $y$ and $z$ (the leftovers).
  Note that the remaining context $\Gamma$ is preserved unused as a leftover.
  This splitting $\opsquared{x}{y}{z}$ is as per the usage algebra provided by the developer for the index $idx$.
  We use the \emph{computability} of the monoidal relation to alleviate the user from the proof burden $\opsquared{x}{y}{z}$.
  The inductive case $\suc$ appends the type $t'$ to the typing context, and the usage annotation $x'$ to both the input and output usage contexts. 
\end{nidefinition}

@@

\begin{nidefinition}[\textsc{Types}: typing processes]
  \label{def:types}

@@

  The inaction process in rule $\PO$ does not change usage annotations.
  The scope restriction in rule $\new$ expects three arguments: the type $t$ of data being transmitted; the usage annotation $x$ of what is being transmitted; and the multiplicity $y$ given to the channel itself.
  This multiplicity $y$ is used for both input and output, so that they are balanced.
  The continuation process $P$ is provided with the new channel with usage annotation $y \comma y$, which it must completely exhaust.
%
  The input process in rule $\recv{}{}$ requires a channel $chan_i$ at index $i$ with usage $\li$ available, such that data with type $t$ and usage $x$ can be sent over it.
  Note that the index $i$ is used in the syntax of the typed process.
  We use the leftovers $\Xi$ to type the continuation process, which is also provided with the received element --- of type $t$ and multiplicity $x$ --- at index $\constr{0}$.
  The received element $x$ must be completely exhausted by the continuation process.
%
  Similarly to input, the output process in rule $\send{}{}$ requires a channel $chan_i$ at index $i$ with usage $\lo$ available, such that data with type $t$ and usage $x$ can be sent over it.
  We use the leftover context $\Delta$ to type the transmitted data, which needs an element $loc_j$ at index $j$ with type $t$ and usage $x$, as per the type of the channel $chan_i$.
  The leftovers $\Xi$ are used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.
%
  Parallel composition in rule $\comp{}{}$ uses the leftovers of the left-hand process to type the right-hand process.
  By keeping track in the typing derivation of $P$ of the resources $P$ uses, we can use them to type $Q$ and save the user from having to provide a top-down proof of context split.
\end{nidefinition}

@@

\begin{nitheorem}[Framing]
  \label{thm:framing}
  Let $P$ be well typed in $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$.
  Let $\Delta$ be such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Let $\Gamma_r$ and $\Xi_r$ be arbitrary contexts such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}

\paragraph*{Weakening}
This property states that inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.
Let $\func{ins}_i$ insert an element into a context at position $i$ --- for simplicity, we use it both to insert types into typing contexts and usage annotations into usage contexts.
\begin{nitheorem}[Weakening]
  \label{thm:weakening}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, lifting every variable greater than or equal to $i$ in $P$ is well typed in
  $\types{\func{ins}_i \; t \; \gamma}{\func{ins}_i \; x \; \Gamma}{\func{lift}_i \; P}{\func{ins}_i \; x \; \Xi}$.
\end{nitheorem}

\paragraph*{Strengthening}
This property states that removing an unused variable preserves the well-typedness of a process.
Let $\func{del}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
\begin{nitheorem}[Strengthening]
  \label{thm:strengthening} 
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable not in $P$, such that $uP \; : \; \Unused_i \; P$.
  Then lowering every variable greater than $i$ in $P$ is well typed in $\types{\func{del}_i \; \gamma}{\func{del}_i \; \Gamma}{\func{lower}_i \; P \; uP}{\func{del}_i \; \Xi}$.
\end{nitheorem}

\paragraph*{Exchange}
This property states that the exchange of two variables preserves the well-typedness of a process.
We extend $\func{exchange}_i$  introduced in \autoref{structural-congruence} to exchange types in typing contexts and usage annotations in usage contexts.
\begin{nitheorem}[Exchange]
  \label{thm:exchange}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, $\types{\func{exchange}_i \; \gamma}{\func{exchange}_i \; \Gamma}{\allowbreak \func{exchange}_i \; P}{\func{exchange}_i \; \Xi}$.
\end{nitheorem}

@@

\begin{nilemma}
  \label{lm:lower-lift}
  The function $\func{lower}_i \; P \; uP$ has an inverse $\func{lift}_i \; P$ that increments every $\textsc{Var}$ greater than or equal to $i$, such that $\func{lift}_i \; (\func{lower}_i \; P \; uP) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nilemma}
  \label{lm:exchange-exchange}
  The function $\func{exchange}_i \; P$ is its own inverse: $\func{exchange}_i \; (\func{exchange}_i \; P) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nilemma}
  \label{lm:types-unused}
  For all well-typed processes $\types{\gamma}{\Gamma}{P}{\Xi}$, if the variable $i$ is unused within $P$, then $\Gamma$ at $i$ is equivalent to $\Xi$ at $i$.
\end{nilemma}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{nitheorem}[Subject congruence]
  \label{thm:subject-congruence}
  If $P \eq{} Q$ and $\types{\gamma}{\Gamma}{P}{\Xi}$, then $\types{\gamma}{\Gamma}{Q}{\Xi}$.
\end{nitheorem}

@@

\begin{nitheorem}[Generalised substitution]
  \label{thm:substitution-generalization}
  Let process $P$ be well typed in $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  Then, there must exist some $\Gamma$, $\Psi$, $\Gamma_j$ and $\Psi_j$ such that:
  \begin{multicols}{2}
  \begin{itemize}
    \item $\contains{\gamma}{\Gamma_i}{i}{t}{m}{\Gamma}$
    \item $\contains{\gamma}{\Gamma_j}{j}{t}{m}{\Gamma}$
    \item $\contains{\gamma}{\Psi_i}{i}{t}{n}{\Psi}$
    \item $\contains{\gamma}{\Psi_j}{j}{t}{n}{\Psi}$
  \end{itemize}
  \end{multicols}
  Let $\Gamma$ and $\Psi$ be such that $\opctx{\Gamma}{\Delta}{\Psi}$ for some $\Delta$.
  Let $\Delta$ at position $i$ have usage $\lz$, meaning all consumption from $m$ to $n$ must happen in $P$.
  Then substituting $i$ to $j$ in $P$ will be well typed in $\types{\gamma}{\Gamma_j}{\subst{P}{j}{i}}{\Psi_j}$.
\end{nitheorem}

@@

\begin{nitheorem}[Substitution]
  \label{thm:substitution}
  Let process $P$ be well typed in $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \lz}$.
  Let $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute the variable references to $\constr{0}$ in $P$ with $\suc j$ so that the result is well typed in $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{0}}}{\Xi \comma m}$.
\end{nitheorem}
\begin{proof}[Proof]
  For $\contains{\gamma}{\Gamma}{j}{t}{m}{\Theta}$ and $\types{\gamma \comma t}{\Theta \comma m}{P}{\Xi \comma \lz}$ for some $\Theta$, we use framing to derive them.
  Then, we use these to apply \autoref{thm:substitution-generalization}.
\end{proof}

@@

\begin{nilemma}
  \label{lm:comm-capable}
  Every input usage context $\Gamma$ of a well-typed process $\types{\gamma}{\Gamma}{P}{\Delta}$ that reduces by communicating on a channel external (that is, $P \reduce{\constr{external} \; i} Q$ for some $Q$) has a multiplicity of at least $\lio$ at index $i$.
\end{nilemma}

@@

\begin{nitheorem}[Subject reduction]
  \label{thm:subject-reduction}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$ and reduce such that $P \reduce{c} Q$.
  \begin{itemize}
    \item If $c$ is $\constr{internal}$, then $\types{\gamma}{\Gamma}{Q}{\Xi}$.
    \item If $c$ is $\constr{external} \; i$ and $\containsusage{\Gamma}{i}{\lio}{\Delta}$, then $\types{\gamma}{\Delta}{Q}{\Xi}$.
  \end{itemize}
\end{nitheorem}
